---
title: "Investigating MS Lesion Scans"
author: "Ying Jin"
date: "`r Sys.Date()`"
output: 
  html_document:
    self_contained: true
    number_sections: true
    code_folding: show
    toc: true
    toc_depth: 3
    toc_float: true
    font: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(neurobase) # https://cran.r-project.org/web/packages/neurobase/index.html
library(tidyverse)
library(here)
```


```{r load_data}
### File paths ####
# list.files("DataRaw")
paths <- read.csv(here("DataRaw/WM_lesion_QC_result.csv"))
# 82 scans
# 8 subjects
# 4 sites
# two scans each sites

#### Read processed image and segmentation ####
paths <- paths %>% 
  separate(., col=flair_files, 
           into = c(rep(NA, 8), "registration", "flair", "name1"),
           sep = "/") %>% 
  separate(., col=mimosa_files, 
           into = c(rep(NA, 8), "mimosa", "name2"),
           sep = "/") 

paths <- paths %>%
  mutate(subject = gsub("-", "", subject)) %>%
  mutate(
  file_path = paste0("DataRaw/processed/data/sub-", subject, "/ses-", session, 
                     "/", registration, "/", flair, "/", name1),
  seg_path =  paste0("DataRaw/processed/data/sub-", subject, "/ses-", session, 
                     "/", mimosa, "/", name2)
) 

# let's use one subject as an example 
paths_01001 <- paths %>% filter(subject=="01001") 

# reading in and display
list_mat <- list()
list_img <- list()
list_seg_img <- list()
list_seg_mat <- list()
for(p in seq_along(paths_01001$file_path)){
  
  # image
  img_p <- readNIfTI(paths_01001$file_path[p], reorient = F)
  list_img[[p]] <- img_p
  list_mat[[p]] <- img_p@.Data
  # segmentation
  seg_p <- readNIfTI(paths_01001$seg_path[p], reorient = F)
  list_seg_img[[p]] <- seg_p
  list_seg_mat[[p]] <- seg_p@.Data
}
```


# Example of subject 01-001

## Whole images

Let's start by looking at the entire images. The lesion is only a very small part of it.

```{r BWH}
ortho2(x=list_img[[1]], y=list_seg_img[[1]], crosshair=FALSE,  
       text = paths_01001$session[1], 
       text.x=40, text.y = 10, text.cex = 1)
ortho2(x=list_img[[2]], y=list_seg_img[[2]], crosshair=FALSE,
       text = paths_01001$session[2], 
       text.x=40, text.y = 10, text.cex = 1)
```


```{r JHU}
ortho2(x=list_img[[3]], y=list_seg_img[[3]], crosshair=FALSE,  
       text = paths_01001$session[3], 
       text.x=40, text.y = 10, text.cex = 1)
ortho2(x=list_img[[4]], y=list_seg_img[[4]], crosshair=FALSE,
       text = paths_01001$session[4], 
       text.x=40, text.y = 10, text.cex = 1)
```


```{r NIH}
ortho2(x=list_img[[5]], y=list_seg_img[[5]], crosshair=FALSE,  
       text = paths_01001$session[5], 
       text.x=40, text.y = 10, text.cex = 1)
ortho2(x=list_img[[6]], y=list_seg_img[[6]], crosshair=FALSE,
       text = paths_01001$session[6], 
       text.x=40, text.y = 10, text.cex = 1)
```


```{r Penn}
ortho2(x=list_img[[7]], y=list_seg_img[[7]], crosshair=FALSE,  
       text = paths_01001$session[7], 
       text.x=40, text.y = 10, text.cex = 1)
ortho2(x=list_img[[8]], y=list_seg_img[[8]], crosshair=FALSE,
       text = paths_01001$session[8], 
       text.x=40, text.y = 10, text.cex = 1)
```

## Preliminary analysis 

I am not exactly sure what to do here, since we do not have two measures for each pixel. Also, the size/location of lesion would change across scans and site. There is no one-to-one relationship here. 

If the goal is to generate image, then it would not be a problem. But if the goal is to generate WMF and QSM, we would at least need to know how these values are calculated. 

I think we would need to focust on the lesion part (is that correct?) and also that would reduce the data by quite a bit. So I hope to start there. 

I think I can start by doing things with the whole image. At least there would be one-to-one relationship across the scans at the same site. 


```{r Scatterplots}
locs <- c("BWH", "Hopkins", "NIH", "Penn")

for(i in 1:4){

    pscan <- data.frame(
        scan1 = as.vector(list_mat[[2*i-1]]),
        scan2 = as.vector(list_mat[[2*i]]),
        lesion1 = as.vector(list_seg_mat[[2*i-1]]),
        lesion2 = as.vector(list_seg_mat[[2*i]])
      ) %>%
        mutate(lesion = paste0(lesion1, lesion2)) %>%
        mutate(lesion = factor(lesion, levels = c("00", "10", "01", "11"),
                               labels = c("", "Scan1", "Scan2", "Both"))) %>%
        ggplot()+
        geom_point(aes(x=scan1, y=scan2, col = lesion), alpha = 0.5, size = 0.2)+
        geom_smooth(aes(x=scan1, y=scan2), method = "lm", formula = y~x)+
        ggpubr::stat_cor(aes(x=scan1, y=scan2), method="spearman")+
        scale_color_brewer(type = "div", palette = "BrBG", direction = -1)+
        labs(title = locs[i], x = "Scan1", y="Scan2")
    print(pscan)
  
}
```



